Assessing \Htwo{} and \COtwo{} Network Strategies in Europe

### 2023-12-14 report/submissions/journal/paper/paper.tex
address TODOs until 3.3

Recently, policymakers and industry in Europe have started developing hydrogen and carbon management strategies~\cite{GermanyDevelopingStrategy2023,CarbonManagementStrategie}, planning infrastructure components~\cite{CONetz}, and committing on the first carbon utilization projects~\cite{EFuelsPilotPlant2022,OrstedAssumesFull,GROUNDBREAKINGEFUELPRODUCTION,DLREfuelsDLR}. With the European Union's goal of achieving climate neutrality by 2050, a wide range of programs, funding models and initiatives have been established with dedicated instruments supporting both the ramp-up of the hydrogen and carbon economy~\cite{eu2023netzero,europeangreendeal,europeaninnovationfund}.
Collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier. Some natural gas pipelines are already repurposed to transport hydrogen~\cite{RohrFreiFuer}. Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synthetic fuels on the one hand and offer \COtwo{} offtake on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \COtwo{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential~\cite{ToolsGreenTransition} of 1.7 Gt carbon annually from point sources, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it's important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily for enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

@@

In summary, both network models prefer CC from point sources distributed across the continent as well as electrolysis concentrated in regions with abundant renewable resources. The two network approaches reveal fundamentally different strategies to integrate these low-cost resources into the system: The \carbonmodel{} transports and sequesters low-cost carbon from distributed point sources and co-locates CU with low-cost electrolysis and DAC. The \hydrogenmodel{} distributes CU alongside with carbon point sources to which low-cost hydrogen from regions with abundant renewable resources is transported.
%
% However, differences between the two models are notable for the placement of CU, CS and DAC due to their distinct \COtwo{} and \Htwo{} price patterns. CU locations are more distributed in the \modH{}, while they cluster in a few valleys of low-cost \Htwo{} in the \modCO{}. CS co-locates in the regions with the lowest \COtwo{} costs (among those with offshore sequestration potential), but these regions are very different between the models. DAC is placed in regions with the best renewable resources in both models, but it is used for only CU in the \modCO{} and for both CU and CS in the \modH{}.
Finally, the system cost advantage of the \modH{} stems from being able to supply the exogenous hydrogen demands in Central Europe with low-cost hydrogen imports, while the \modCO{} is forced to deploy electrolysis in regions with poor renewables resources as well as additional gas infrastructure in Central Europe.

@@

% TODO: smoother transition, incl. end of 3.2
% TODO: add network investment costs plot to Appendix. Mention investment H2 > CO2. (The investment in the hydrogen network is four times the investment in the carbon network.)
The \modHybrid{} combines the advantages of hydrogen and carbon networks. Fig.~\ref{fig:cost_bar} shows that the \modHybrid{} achieves a cost reduction of \label{}5.2\% (\label{}41 billion euros) compared to the \modBase{}, which is lower than the cost reductions of the \COgrid{} and \modH{}. Fig.~\ref{fig:capacity_maps} shows the average production, consumption and transport of carbon and hydrogen in the \modHybrid{}. While the topology of the \Htwo{} network roughly corresponds to that of the \modH{}, the \COtwo{} network plays more of a complementary role. In this section, we explore the advantages and disadvantages compared to the other models.

Similar to the \modH{}, the \modHybrid{} transports hydrogen from concentrated electrolysis sites in Western regions (Iberian Peninsula and British Isles) to distributed CU production sites across the continent. For the \COtwo{} network, this model employs short and direct pipeline routes, typically spanning one or two segments, to transport carbon from inland CC point sources to perform CS at nearby sequestration sites (e.g., Northern Germany, Italy and Greece). Additionally, the \COtwo{} network utilizes single-segment network sections to equalize local variations in carbon demand and supply to produce CU, thereby also further converging carbon prices.

To understand the dynamics of the \modHybrid{} model, we observe that the model locates CU everywere, except for a few coastal regions that perform CS, or that supply their locally captured carbon to a neigbouring sequestration region. The model uses the \COtwo{} network to transport carbon over short distances to the CS sites, and the \Htwo{} network to transport hydrogen over longer distances from the lowest-cost hydrogen regions to CU sites across the continent. DAC is deployed to supply carbon for the remaining CU needed by the model (mostly in Southern Spain).

Despite only a further reduction of \label{}0.4 percentage points (\label{}3~bn€) in total system costs compared to the \modH{}, the \modHybrid{} brings several advantages: (1) The \modHybrid{} uses both \COtwo{} and \Htwo{} networks to build a more robust energy system; (2) it relies even less on DAC than the other models, which as a technology may be costlier than anticipated in the cost projections; and (3) it requires a smaller land use footprint for wind and solar power due to lower DAC deployment. On the other hand, the \modHybrid{} needs slightly more biomass, equipped with carbon capture, to achieve its cost advantage in comparison to the \modH{}.

@@

To supply carbon needed for CS and CU, the system can choose to deploy carbon capture (CC) technologies at various point sources of fossil or biogenic origin (see below), or through DAC facilities.
The concept of a merit order for captured carbon plays a pivotal role in optimizing the deployment of carbon capture technologies based on their economic feasibility.
This merit order ranks all CC technologies according to their relative costs to capture an additional marginal ton. The cost value for a technology depends on the exact system configuration and may vary by scenario and even by region, since many technologies produce other (primary) outputs such as electricity and heat. The costs to capture an additional marginal ton of carbon may also be different from the average cost of capture.
At the lower end of the cost spectrum, CC technologies applied to process emissions, such as those from cement, offer a cost-effective starting point.
Following this, biomass combined heat and power (CHP) systems provide a dual benefit of energy production and carbon capture.
Moving up the scale, gas used in industrial applications and biomass employed for industrial processes represent more costly yet viable options to capture carbon.
DAC, an emerging technology capable of extracting \COtwo{} directly from the atmosphere, stands at a higher cost level due to its current high capital costs and energy demand.
Finally, based on our model results, biogas upgrading, a process that refines biogas to natural gas quality, incurs the highest costs in the merit order per marginal ton of captured \COtwo{}.
Biogas input is a high-cost fuel which based on the endogenous modeling decisions is not used in large quantities.
To the extent that biogas-to-gas facilities are used by the model to supply additional (carbon-neutral) gas as fuel, adding CC infrastructure incurs low costs (and thus ranges on the left-hand side of the merit order curve).
However, the price of capturing additional \COtwo{} from biogas-to-gas upgrading is high because a substantial amount of the biogas fuel costs factors into the marginal cost of captured \COtwo{}.
This merit order framework is crucial for strategically deploying necessary carbon capture solutions while balancing economic considerations.
If we consider the spatial aspect of distributed carbon capture potentials, the \COtwo{} network can exploit comparative cost advantages between CO2 bidding zones to utilize the lowest-cost CC facilities across the continent.
We assume a capture rate of 90\% for CC on process emissions, SMR, biogas-to-gas, as well as gas and biomass used in industry, and 95\% for CC on biomass and gas CHPs.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar_transmission.png}
    \caption{Annual transmission system cost for the different models in the net-zero emission scenario.}
    \label{fig:cost_bar_transmission}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth]{comparison/net-negative-0.1/figures/90_nodes/cost_bar_transmission.png}
    \caption{Annual transmission system cost for the different models in the net-negative emissions scenario.}
    \label{fig:cost_bar_transmission_nn}
\end{figure}

@@

\begin{figure*}[ht!]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{net-negative-0.1/co2-only/figures/90_nodes/balance_map_carbon.png}
        \label{fig:capacity_map_carbon_co2_nn}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{net-negative-0.1/co2-only/figures/90_nodes/balance_map_hydrogen.png}
        \label{fig:capacity_map_hydrogen_co2_nn}
    \end{subfigure}
    \caption{Optimal operation, flows and prices of the carbon (left) and hydrogen (right) sectors for the \carbonmodel{} in the net negative scenario. For each region, upper semicircles show the average production per technology, lower semicircles the consumption, and colors the average marginal prices. Carbon Sequestration offshore are drawn as full circles. Lines and arrows show the interregional transportation. The \carbongrid{} reveals a fundamentally different layout than in the Net-Zero scenario.
    }
    \label{fig:capacity_maps_co2_nn}
\end{figure*}

\begin{figure*}[ht!]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{net-negative-0.1/h2-only/figures/90_nodes/balance_map_carbon.png}
        \label{fig:capacity_map_carbon_h2_nn}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{net-negative-0.1/h2-only/figures/90_nodes/balance_map_hydrogen.png}
        \label{fig:capacity_map_hydrogen_h2_nn}
    \end{subfigure}
    \caption{Optimal operation, flows and prices of the carbon (left) and hydrogen (right) sectors for the \hydrogenmodel{} in the net negative scenario. For each region, upper semicircles show the average production per technology, lower semicircles the consumption, and colors the average marginal prices. Carbon Sequestration offshore are drawn as full circles. Lines and arrows show the interregional transportation. With the tightened emission target, the \hydrogengrid{} expands the \hydrogengrid{} layout of the Net-Zero scenario model results while increasing DAC facilities at the coast.
    }
    \label{fig:capacity_maps_h2_nn}
\end{figure*}

\begin{figure*}[ht!]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{net-negative-0.1/full/figures/90_nodes/balance_map_carbon.png}
        \label{fig:capacity_map_carbon_full_nn}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{net-negative-0.1/full/figures/90_nodes/balance_map_hydrogen.png}
        \label{fig:capacity_map_hydrogen_full_nn}
    \end{subfigure}
    \caption{Optimal operation, flows and prices of the carbon (left) and hydrogen (right) sectors for the \hydrogenmodel{} in the net negative scenario. For each region, upper semicircles show the average production per technology, lower semicircles the consumption, and colors the average marginal prices. Carbon Sequestration offshore are drawn as full circles. Lines and arrows show the interregional transportation. With the tightened emission target, the \hybridmodel{} decreases the \hydrogengrid{} layout in comparison to the Net-Zero scenario and increases \carbon{} transport from inland point sources to the sequestration sites.
    }
    \label{fig:capacity_maps_full_nn}
\end{figure*}

>>>>

Recently, policymakers and industry in Europe have started developing hydrogen and carbon management strategies~\cite{GermanyDevelopingStrategy2023,CarbonManagementStrategie}, planning infrastructure components~\cite{CONetz}, and committing on the first carbon utilization projects~\cite{EFuelsPilotPlant2022,OrstedAssumesFull,GROUNDBREAKINGEFUELPRODUCTION,DLREfuelsDLR}. With the European Union's goal of achieving climate neutrality by 2050, a wide range of programs, funding models and initiatives have been established with dedicated instruments supporting both the ramp-up of the hydrogen and carbon economy~\cite{eu2023netzero,europeangreendeal,europeaninnovationfund}.
Collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier. Some natural gas pipelines are already repurposed to transport hydrogen~\cite{RohrFreiFuer}. Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synfuels on the one hand and offer \COtwo{} offtake on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \COtwo{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential~\cite{ToolsGreenTransition} of 1.7 Gt carbon annually from point sources, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it's important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily for enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

@@

In summary, both network models prefer CC from point sources distributed across the continent as well as electrolysis concentrated in regions with abundant renewable resources. The two network approaches reveal fundamentally different strategies to integrate these low-cost resources into the system: The \carbongridmodel{} transports and sequesters low-cost carbon from distributed point sources and co-locates CU with low-cost electrolysis and DAC. The \hydrogengridmodel{} distributes CU alongside with carbon point sources to which low-cost hydrogen from regions with abundant renewable resources is transported.
%
% However, differences between the two models are notable for the placement of CU, CS and DAC due to their distinct \COtwo{} and \Htwo{} price patterns. CU locations are more distributed in the \modH{}, while they cluster in a few valleys of low-cost \Htwo{} in the \modCO{}. CS co-locates in the regions with the lowest \COtwo{} costs (among those with offshore sequestration potential), but these regions are very different between the models. DAC is placed in regions with the best renewable resources in both models, but it is used for only CU in the \modCO{} and for both CU and CS in the \modH{}.
Finally, the system cost advantage of the \modH{} stems from being able to supply the exogenous hydrogen demands in Central Europe with low-cost hydrogen imports, while the \modCO{} is forced to deploy electrolysis in regions with poor renewables resources as well as additional gas infrastructure in Central Europe.

@@

% TODO: smoother transition, incl. end of 3.2
% TODO: inlcude in opener: both networks, lower system costs; motivation: what is the advantage vs both others? other advantages (more resilient)/disadvantages (not focusing on one grid)
The \modHybrid{} combines the advantages of hydrogen and carbon networks.
Fig.~\ref{fig:capacity_maps} shows the average production, consumption and transport of the two networks in the \modHybrid{}.

While the topology of the \Htwo{} network roughly corresponds to that of the \modH{}, the \COtwo{} network plays more of a complementary role.

% TODO: mention investment difference in network expansion with appendix plot (The investment in the hydrogen network is four times the investment in the carbon network.)
% TODO: rewrite: mostly short (1-2 regions) linear (no cluster/mashed network) routes towards sequestration sites (for purpose of seq.); short 1-region transports to bridge local carbon price diffs, for purpose of CU.
Similar to the \modH{}, the \modHybrid{} takes advantage of a large hydrogen network that enables hydrogen to be transported from centralized production sites in Western regions such as Spain to decentralized CU production sites close to CC point sources across the continent.
The carbon network supplements this with smaller clusters of networks. In addition to a larger \COtwo{} network cluster in Central Europe, which transports carbon to the North Sea, the system is building three large linear carbon routes. These transport carbon from Romania and Bulgaria to Greece, from northern Italy to central Italy and from Hungary, the Czech Republic and Poland to the Baltic Sea. These enable the system to transport carbon available at low cost from inland biomass sources and sequester it (see Fig.~\ref{fig:captureshare_line}). This reduces the need for DAC in the system and the sequestration sites are shifted to the end points of the carbon networks. Second-order effects can be seen in the shift in FT production compared to the \modH{}. Regions that have both good renewable resources and sequestration potential could now forego sequestration and use local process emissions for FT production, as can be observed in Ireland. A notable aspect of the hybrid model is the lack of overlap in the hydrogen and carbon network topologies. This finding is tightly linked to the fact of competing functionalities to offtake carbon emissions.
A more detailed of differences between the \modH{} and the \modHybrid{} can be found in Fig.~\ref{fig:cost_map_difference}.

% TODO: add interpretative paragraph: no CU where low-price ely, H2 is imported.
% Both networks available to smooth prices; CS draws in CC from adjacent regions; CU everywhere locally, except the CS sites; DAC centered on (very few) regions with optimal RES (and Italy/Greece for methanation)
% compare to H2 Grid model: küstennahe regionen mit longest distance to good RES(?) anzapfen, dort kein CU, CU jewtzt im Hinterland

Despite only a further reduction of \label{}0.4 percentage points (\label{}3~bn€) in total system costs compared to the \modH{}, the \modHybrid{} brings several advantages. These include a broader range of technologies contributing to its robustness, reduced reliance on DAC, which may be costlier than anticipated in the cost projections, and less land use for wind and solar due to the decreased necessity for DAC. The increase used of biomass with carbon capture, transport and sequestration leads to slight cost advantages in comparison to the \modH{}.

@@

To supply carbon needed for CS and CU, the system can choose to deploy carbon capture (CC) technologies at various point sources of fossil or biogenic origin (see below), or through DAC facilities.
The concept of a merit order for captured carbon plays a pivotal role in optimizing the deployment of carbon capture technologies based on their economic feasibility.
This merit order ranks all CC technologies according to their relative costs to capture an additional marginal ton\footnote[2]{The cost value for a technology depends on the exact system configuration and may vary by scenario and even by region, since many technologies produce other (primary) outputs such as electricity and heat. The costs to capture an additional marginal ton of carbon may also be different from the average cost of capture.}
At the lower end of the cost spectrum, CC technologies applied to process emissions, such as those from cement, offer a cost-effective starting point.
Following this, biomass combined heat and power (CHP) systems provide a dual benefit of energy production and carbon capture.
Moving up the scale, gas used in industrial applications and biomass employed for industrial processes represent more costly yet viable options to capture carbon.
DAC, an emerging technology capable of extracting \COtwo{} directly from the atmosphere, stands at a higher cost level due to its current high capital costs and energy demand.
Finally, based on our model results, biogas upgrading, a process that refines biogas to natural gas quality, incurs the highest costs in the merit order per marginal ton of captured \COtwo{}.
Biogas input is a high-cost fuel which based on the endogenous modeling decisions is not used in large quantities.
To the extent that biogas-to-gas facilities are used by the model to supply additional (carbon-neutral) gas as fuel, adding CC infrastructure incurs low costs (and thus ranges on the left-hand side of the merit order curve).
However, the price of capturing additional \COtwo{} from biogas-to-gas upgrading is high because a substantial amount of the biogas fuel costs factors into the marginal cost of captured \COtwo{}.
This merit order framework is crucial for strategically deploying necessary carbon capture solutions while balancing economic considerations.
If we consider the spatial aspect of distributed carbon capture potentials, the \COtwo{} network can exploit comparative cost advantages between CO2 bidding zones to utilize the lowest-cost CC facilities across the continent.
We assume a capture rate of 90\% for CC on process emissions, SMR, biogas-to-gas, as well as gas and biomass used in industry, and 95\% for CC on biomass and gas CHPs.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth]{comparison/emission-reduction-full/figures/90_nodes/cost_bar_transmission.png}
    \caption{Annual transmission system cost as a function of the net carbon removal scenarios considered in the study.}
    \label{fig:cost_bar_transmission}
\end{figure}

@@
